Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-03 paper/paper.tex
Paper chapters 1,2 ready, 3 nearly

To address these issues we looked at \emph{parameterized random networks}, or
in the language of genetics, at different \emph{network genotypes}. These
genotypes contain parametric information about the networks (\emph{phenotypes}) to be created. 
Different phenotypes with the same genotype may thus
have strong quantitative differences, but they share structural qualities
whose fitness for a preassigned task we wanted to analyze and improve.

@@

\section{Concepts and Model}
\paragraph{Network Model}

MAYBE WE SHOULD MOTIVATE FORCE HERE, WHICH THEN NATURALLY LEADS TO THE CHOICE OF NETWORK BELOW.

We chose the same rate based network architecture
that is used in the original FORCE publication \citep{FORCE} (denoted as
architecture A) and also in \citep{RM}, where our notation mainly stems
from. So we look at a network $\mathcal{N}$ with $N$ internal neurons and
states $x_i$ for $1 \le i \le N$ that loosely represent the neurons'
membrane potential. The firing rate $r_i$ of the $i$-th internal neuron is
given by $r_i = \tanh x_i$.  Further, the network contains $R$ readout
neurons with states $z_i$ for $1 \le i \le R$, where
\begin{equation} 
    z_i = \sum_{j=1}^{N}\omega\Ix{read}_{ij}\,r_j~.
\end{equation}
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}. We in fact enabled
inputs in our code base, but did not use them for the main results.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}_{ij}$ and the feedback synapse weights
$\omega\Ix{fb}_{ij}$. One can see that the neurons exponentially
decay if they are not presented stimulus by other neurons. 
We solved this system of differential equations by
a simple Newton method with integration time steps $dt = 0.01$, which was
also used in \citep{FORCE}. (How we constructed $\omega\Ix{rec}$ and
$\omega\Ix{fb}$ will be discussed in \emph{Parametric Random Networks})

\paragraph{Learning} Learning took place by stepwisely adapting the weights
$\omega\Ix{read}$ so that the dynamics $t\mapsto r_i(t)$ of the readout
rates should eventually match a given periodic input pattern $t\mapsto
a_i(t)$ as accurately as possible. To update $\omega\Ix{read}$ we chose
a variant of the FORCE algorithm \citep{FORCE} that rapidly modifies the
feedback loop to keep the errors $|r_i(t) - a_i(t)|$ small from the
beginning on.

While the authors of \citep{FORCE} update the weights $\omega\Ix{read}$ in
fixed intervals (every second time step), we used an adaptive mechanism to
determine how often learning should occur. This was initially introduced
for performance reasons, as the single learning steps proved to be the main
computational bottleneck, but we also found that the long time network
dynamics tended to be slightly more accurate for adaptive learning steps.

\paragraph{Parametric Random Networks -- Genotypes} Since we are not
interested in single network properties but rather want to find out which
structural elements of networks are important for learning success, we
introduced \emph{parametric random networks}, or \emph{network genotypes}.
Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
which concrete networks are created when $G$ is expressed in a single
phenotype. Technically $G$ may be understood as a stochastic network
generator that returns a network $\mathcal{N}=G(\omega)$ for every seed
value $\omega$.
The genotypes' parameters $\theta\in\Theta$ were either integer or floating
point values and can be classified as affecting either (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Important examples for (1) are the networks size $N$, the
occupation probability $p$ (for random Erdös-Renyi networks), the
neighborhood range $k$ (for ring topologies), or the number $l$ of layers
(for feed-forward networks). We furthermore introduced ratio parameters
that allowed us to interpolate between different topologies.  Parameters of
type (2) were the feedback strength $\texttt{feedback}$ (scaling of
$\omega\Ix{fb}$) and the gain value \gain (scaling of
$\omega\Ix{rec}$ by ).

\paragraph{Fitness} For a given genotype $G$ and a given challenge $C$, the
fitness $f_C(G)$ of $G$ for $C$ is taken to be the expected value of
the success rates of $G$'s phenotypes when solving the tasks of the
challenge $C$. In our simulations, $f_C(G)$ was of course only
approximated; this was usually done by taking 20 to 100 sample phenotypes
and sample tasks, where the exact number depended on the standard deviation
of the previously drawn samples' success rates.

\section{Methods and Implementation}

\paragraph{Tasks}
For the learning to work properly, the problems presented to the
networks have (for the most part) to be solvable by FORCE. While
it was shown \citep{FORCE} that large recurrent neural networks are capable of
learning very complex functions, the optimization requires a large number
of learnings. Therefore, the networks have to be restricted to being quite
small ($\approx 100$ neurons) and are consequently less powerful.

To get a simple criterion to estimate whether it might be possible to learn a given function 
using FORCE with 100 neurons, we look at a single sinusoidal of frequency $\omega$ and 
determine the range of $\omega$ in which an (unoptimized) network (i.e. \gain, 
\feedback, \ERprob) is able to learn the function (see fig. \ref{}).

@@

\paragraph{Success}
In order to optimize the function reconstruction of a force network,
a measure $f(r', r)$ for the concordance of the target function
$r'(t)$ and the network readout $r(t)$ is necessary. Simply using the
difference $f\ix{d}(r, r') = \left\|r-r'\right\|_2$ has the
drawback that small phase differences could produce huge differences in
quality, even though $r'$ has the exact same shape as $r$.

A measure that is time-shift independent and maximized for functions of identical shape is the 
maximum of the cross correlation 
\begin{align}
    f\ix{c}(r, r') = \max_t \frac{\left<r, r'(\cdot - t)\right>}{\sqrt{\left< r, r\right> \left<r', r' \right>}}~.
\end{align}
Algorithmically, the signals are split into overlapping chunks which are correlated separately. 
This saves computational power and prevents minute differences in frequency to accumulate into 
a huge phase shift, i.e. the phase only has to remain relatively
constant over the timescale of a single chunk (about 10 seconds of simulated time).

This measure becomes problematic for functions that are very close to zero for a longer time 
(order of chunk size), but since we are going to work with superpositions of sinusoidals of 
different frequencies, this drawback is unproblematic.

\paragraph{Genetic Optimization}
Due to the stochastic nature of the fitness function $F_C(G)$ of a generator $G$, and the fact that single evaluations
of $F_C$ are computationally expensive, thus preventing increasing the number of samples such that the estimation $E(F_C)$
becomes sufficiently smooth, typical optimization methods like gradient descent or simulated annealing are not usable.

 On approach would be to use techniques from Monte Carlo Integration, e.g. stratified or importance sampling, to aid in the
evaluation of $E(F_C)$, but since $G(\lambda)$ depends non-continuously on $\lambda$, it can only be applied to the dependence 
on the challenge $C$. This would mitigate the effects, but not solve the problem.

Therefore, we optimized using a genetic algorithm, which is inherently capable of coping with the stochastic fitness function.

In each iteration of the algorithm, a generation (set) of $N$ individuals (Generators) $G_t = \left\{ g_i \right\}$ is converted into
a new one $G_{t+1}$ according to the following scheme:
\begin{description}
 \item[Evaluation] The fitness $f_i = F_C(g_i)$ of each individual is measured using at least 20 samples (and adaptively more if the $f_i$ are spaced closely, to at msot 100).
 \item[Selection] All individuals are grouped into disjunct pairs for which we check which one has the higher $f$. This is repeated $K=10$ times, counting the wins of each individual. Based thereon we use the $p=50\%$ best individuals $S$ to build the next generation.

By using pair comparisons instead of simple ordering according to $f$, we allow weaker individuals to continue into the next generation albeit with low probability), keeping the gene pool more diverse.

\item[Reproduction]
We generate the next generation $G_{t+1}$ by first generating $2N$ candidates by \emph{mutating} or \emph{recombining} elements of $S$, i.e. either changing a single parameter randomly or taking two individuals and randomly choose for each parameter from which parent to take the value. 
We then perform a single sample fitness test and remove all networks that fail, until only $N$ networks are left which form $G_{t+1}$. 
This way, the most unpromising elements of the new generation are removed before the expensive fitness evaluation. 
\end{description}

@@

\section{Results}

>>>>

To address these issues we looked at parameterized random networks, or in
the language of genetics, at different \emph{network genotypes}. These
genotypes contain parametric information about the networks
(\emph{phenotypes}) to be created. 
Different phenotypes with the same genotype may thus have strong
quantitative differences, but they share structural qualities whose fitness
for a preassigned range of tasks we wanted to analyze and improve.

@@

We first lay out the main aspects of our model in a 
partially abstract fashion in the next section.
%rather high-level fashion. 
More details on the actual implementation are then provided in the third
section (or ultimately in the published source
code / supplementary material).
Finally, some of the obtained results are presented and discussed.

\section{Concepts and Model}
\label{sec:concepts}
\paragraph{Network Model} 
%MAYBE WE SHOULD MOTIVATE FORCE HERE, WHICH THEN NATURALLY LEADS TO THE CHOICE OF NETWORK BELOW.

We chose the same rate based network architecture
that is used in the original FORCE publication \citep{FORCE} (denoted as
architecture A) and also in \citep{RM}, where our notation mainly stems
from. So we look at a network $\mathcal{N}$ with $N$ internal neurons and
states $x_i$ for $1 \le i \le N$ that loosely represent the neurons'
membrane potential. The firing rate $r_i$ of the $i$-th internal neuron is
given by $r_i = \tanh x_i$.  Furthermore, the network contains $R$ readout
neurons with states $z_i$ for $1 \le i \le R$, where
\begin{equation} 
    z_i = \sum_{j=1}^{N}\omega\Ix{read}_{ij}\,r_j
\end{equation}
with the readout weights $\omega\Ix{read}$.
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}. We in fact enabled
inputs in our code base, but did not use them for the main results.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}_{ij}$ and the feedback synapse weights
$\omega\Ix{fb}_{ij}$. One can see that the neurons exponentially
decay if they are not presented stimulus by other neurons. 
We solved this system of differential equations by
a simple Newton method with integration time steps $dt = 0.01$, which was
also used in \citep{FORCE}. 
%(How we constructed $\omega\Ix{rec}$ and
%$\omega\Ix{fb}$ will be described in \emph{Parametric Random Networks})

\paragraph{Learning} Learning took place by stepwisely adapting the weights
$\omega\Ix{read}$ so that the dynamics $t\mapsto r_i(t)$ of the readout
neurons should eventually match a given periodic target pattern $t\mapsto
r'_i(t)$, called a \emph{task} from now on, as accurately as possible. To
update $\omega\Ix{read}$ we chose a variant of the FORCE algorithm
\citep{FORCE} that rapidly modifies the feedback loop to keep the errors
$|r_i(t) - r'_i(t)|$ small from the beginning on.

While the authors of \citep{FORCE} update the weights $\lambda\Ix{read}$ in
fixed intervals (every second time step), we used an adaptive mechanism to
determine how often learning should occur. This was initially introduced
for performance reasons, as the single learning steps proved to be the main
computational bottleneck, but we also found that the long time network
dynamics tended to be slightly more accurate for adaptive learning steps.

\paragraph{Network Genotypes} Since we are not
interested in single network properties but rather want to find out which
structural elements of networks are important for learning success, we
introduced  \emph{network genotypes}.
Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
which concrete networks are created when $G$ is expressed in a single
phenotype. Technically $G$ may be understood as a stochastic network
generator that returns a network $\mathcal{N}=G(\lambda)$ for every seed
value $\lambda$.
The genotypes' parameters $\theta\in\Theta$ were either integer or floating
point values and can be classified as affecting either (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Important examples for (1) are the networks size $N$, the
occupation probability $p$ (for random Erdös-Renyi networks), the
neighborhood range $k$ (for ring topologies), or the number $l$ of layers
(for feed-forward networks). We furthermore introduced ratio parameters
that allowed us to interpolate between different topologies. Optimized
parameters of type (2) were the feedback strength $\feedback$ (with
$\omega\Ix{fb}\propto \feedback$) and the gain value $\gain$ (with
$\omega\Ix{rec} \propto \gain$).

\paragraph{Fitness} The next step was to define the \emph{fitness} of a given
genotype when confronted with a diverse range of tasks, like waves of
different frequencies. To express this formally, we introduced
\emph{challenges}: For a given seed value $\lambda$, the challenge $C$
yields a task $r' = C(\lambda)$ according to some (parametrized)
distribution. If we now make use of a success function $S(\mathcal{N},
r')\in\{0, 1\}$ that decides, if the network $\mathcal{N}$
has successfully reproduced the desired output $r'$, the fitness $F_C(G)$
of a genotype $G$ for a challenge $C$ may reasonably be defined as the
expectation value of $\lambda \mapsto S(G(\lambda), C(\lambda))$. For fixed
challenges, the main task of optimizing the genotype is now reduced to
finding parameters $\Theta$ that maximize $F_C$.

% IDEA: Maybe put this part at the end of the paper. Most things needed to
% understand the results (in principle) have already been mentioned at this
% point.
\section{Methods and Implementation}
\label{sec:implementation}

\paragraph{Tasks}
For the optimization to work properly, the tasks presented to the networks
must be reasonably chosen; they must be learnable by FORCE (at least in
principle) and must be compatible with technical aspects like the
integration time step.
%have (for the most part) to be solvable by FORCE. 
While it was shown in \citep{FORCE} that large recurrent neural networks are
capable of learning very complex tasks, the optimization requires an
extensive number of learning processes, since every single evaluation of
the fitness $F_C$ needs dozens of tested phenotypes (see below). Therefore,
the networks had to be restricted to being quite small ($\approx 100$
neurons) and are consequently less powerful.

To get a first estimate whether it might be possible to learn a given task
using FORCE with 100 neurons in our network model, we looked at a single
sinusoidal of frequency $\omega$ and determined the frequency range in
which an (unoptimized) network (i.e. \gain, \feedback, \ERprob) is able to
learn the function (see fig. \ref{}).

@@

\paragraph{Success}
As mentioned in \emph{Concepts and Methods}, we need a function $S$
that decides weather a network succeeded or failed to solve a task.
To this end
%In order to optimize the function reconstruction of a force network,
a measure $f(r', r)$ for the concordance of the target function $r'(t)$ and
the network readout $r(t)$ is sought. Though one could use a simple
difference based metric, like the norm $f\ix{2}(r, r')
= \left\|r-r'\right\|_2$, this has the drawback that small phase
mismatches could produce high differences, even though the
shape of $r$ matches the one of $r'$ well.

Instead, a measure that is time-shift independent and maximized for functions of
identical shape is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(r, r') = \max_t \frac{\left<r, r'(\cdot
    - t)\right>}{\sqrt{\left< r, r\right> \left<r', r' \right>}}~.
\end{align}
Algorithmically, the signals $r$ and $r'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
This measure becomes problematic for functions that are very close to zero
for a longer time (order of chunk size), but since we are going to work
with superpositions of sinusoidals of different frequencies, this drawback
is unproblematic.

Using this measure of quality, the success function for a network
$\mathcal{N}$ with readout $r$ for the task $r'$ was taken to be 
\begin{equation}
    S(\mathcal{N}, r') = \begin{cases} 1\quad\mathrm{if}~f\ix{cor}(r, r') \ge 0.95,\\
                                       0\quad\mathrm{if}~f\ix{cor}(r, r') < 0.95~.
                         \end{cases}
\end{equation}
The threshold value of 0.95 was determined empirically, as signals $r$ with
$f(r, r') \ge 0.95$ met our assessment of what we considered a good
reproduction of the target pattern.

\paragraph{Genetic Optimization}
Due to the stochastic nature of the fitness function $F_C(G)$ of a genotype
$G$, a very high number of samples would be required to obtain sufficiently
smooth approximations of $F_C(G)$ for typical optimization methods like
gradient descend or simulated annealing. But as the computational cost for
even one learning process was considerable high, these methods were not
feasible.

%and the fact that single evaluations of $F_C$ are
%computationally very expensive, thus preventing increasing the number of
%samples such that the estimation $E(F_C)$ becomes sufficiently smooth,
%typical optimization methods like gradient descent or simulated annealing
%are not usable.

An alternative approach would be to use techniques from Monte Carlo
integration, e.g.  stratified or importance sampling, to aid in the
evaluation of $F_C(G)$, but since $G(\lambda)$ and $C(\lambda)$ depend
non-continuously on the seed value $\lambda$, this too would not solve the
problem.

Therefore, we optimized $F_C(G)$ using a genetic algorithm, which is inherently
capable of coping with the stochastic fitness function. In each iteration
of the algorithm, a generation $\mathcal{G}_t = \left\{ G_i~|~1\le i\le N\right\}$
of $N$ individuals (Generators) is converted into a new one, $\mathcal{G}_{t+1}$,
according to the following scheme:
\begin{description}
 \item[Evaluation] The fitness $F_i = F_C(G_i)$ of each individual is
     measured using at least 20 samples (and adaptively more if the
     $(F_i)_{1\le i\le N}$ are spaced closely, to at most 100).
 \item[Selection] All individuals are grouped into disjunct pairs for which
     we check which one has the higher fitness. This is repeated $k=10$ times,
     counting the number of wins for each individual. Based thereon we use the
     $p=50\%$ best individuals $\mathcal{G}\Ix{survive}_t$ to build the next
     generation.  By using pair comparisons instead of simple ordering
     according to the fitness, we allow weaker individuals to reach
     the next generation (albeit with low probability), keeping the gene
     pool more diverse.

\item[Reproduction] We obtain the next generation $\mathcal{G}_{t+1}$ by first
    generating $2N$ candidates via \emph{mutating} or \emph{recombining}
    elements of $\mathcal{G}\Ix{survive}_t$. This means we either change
    a single parameter randomly or take two individuals and randomly choose
    for each parameter from which parent to take the value.  We then
    perform a single sample fitness test and remove all networks that fail,
    until only $N$ networks are left; these form $\mathcal{G}_{t+1}$. By
    this approach the most unpromising elements of the new generation are
    removed before the expensive fitness evaluation takes place. 
\end{description}

@@

\section{Results}
Baseline
gain and feedback tests

(feedback shifts possible frequencies to higher values)

success of omega: erdös-renyi, ring
optimizations: ER topology, Ring Topology, Mixing
